Regularised parameter selection in the non-stationary Hawkes process

### 2017-05-08 _paper_output/chapter_sparse_hawkes.tex
fix up some typos

This is analagous to a blind deconvolution problem
\cite{stockham_blind_1975}, adaptive robust regression
\cite{donoho_automatic_1988}, or a sparse inverse problem
\cite{tibshirani_regression_1996}, but considerably more challenging;
While the process model superficially resembles an i.i.d. linear
process, the different noise model and sequential dependence mean that
there are few of the same convenient results; We cannot easily, for
example, choose our our penalty parameters by standard cross validation.
The proposed method uses a heuristically-motivated degrees-of-freedom
regularisation parameter selection, which we test through simulation.
The proposed method works well in practice.

@@

We are given a process \(N:\cc{B}(\bb{R})\mapsto \bb{N}^0\), which
counts the number of arrivals in a given time. Since we are usually
concerned with arrivals over increasing time intervals, we will usually
consider \(N\) to be indexed by time:
\(N(t):=N([0,t))-N((-\infty, t))\).

We represent a realisation of the time series by a (finite by
assumption) sequence of arrival times
\(\v{t}=\{\dots,t_1, t_2,\dots,t_N(T)\}.\) We will be concerned with
estimating its likelihood based on the likelihood of the series on an
interval an interval \([0,T]\). We define \(\v{t}':=\v{t}\cap [0,T]\)
and assume \(\left|\v{t}'\right|<\infty.\)

@@

    We use only the left-continuous filtration
\(\mathcal{F}(\{N(s)\}_{s < t})\) generated by the paths of the process,
which we here write as \(\cc{F}(N(t)).\)

We call it a regular point process if it has an associated
\emph{intensity} process \(\lambda\), such that \(\lambda\) is
predictable with respect to \(\cc{F}(N)\) generated by the process, i.e.
\cite{daley_introduction_2003}

Hereafter we define \(\lambda^*(t):=\lambda(t|\cc{F}(N(t)))\) for
brevity.

Such regular point processes generalise the homogeneous Poisson rate in
the sense that if we choose \(\lambda^*(t)\equiv \lambda_0,\) then it is
the intensity of a homogeneous Poisson rate with fixed intensity
\(\lambda_0\).

    We will use the inter-arrival waiting times \(s_i=t_{i+1}-t_i\) to
calculate the data likelihood, after Rubin \cite{rubin_regular_1972}.

From the definition of the intensity we find \[\begin{aligned}
\frac{\dd}{\dd t}\bb P(t_{i+1}=t|\mathcal{F}(N(t)))=
\end{aligned}\]

TODO: Which bits does Z want proven? Rewrite to start from transition
probability? Probably makes this shorter. Otherwise prove equivalence of
intensity and transition prob DE?

@@

\[
\begin{aligned}
\lambda(t|\mathcal{F}(N(t);\mu,\phi) &= \mu(t) + \phi * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi(t-s)\dd N(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi(t-s)\dd N(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi(t-t_i)
\end{aligned}
\]

For brevity, we write
\(\lambda(\mathcal{F}(N(t));\mu,\phi)=\lambda^*(t)\), but will suppress
explicity dependence on history and parameters..

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_14_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_30_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_30_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_30_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

\texttt{\color{outcolor}Out[{\color{outcolor}27}]:}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_32_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

\texttt{\color{outcolor}Out[{\color{outcolor}28}]:}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_34_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

>>>>

This is analagous to a blind deconvolution problem
\cite{stockham_blind_1975}, adaptive robust regression
\cite{donoho_automatic_1988}, or a sparse inverse problem
\cite{tibshirani_regression_1996}, but considerably more challenging;
While the process model superficially resembles an i.i.d. linear
process, the different noise model and sequential dependence mean that
there are fewer convenient results; We cannot easily, for example,
choose our our penalty parameters by standard cross validation. The
proposed method uses a heuristically-motivated degrees-of-freedom
regularisation parameter selection, which we test through simulation.
The proposed method works well in practice.

@@

We are given a process \(N:\cc{B}(\bb{R})\mapsto \bb{N}^0\), which
counts the number of arrivals in a given time. Since we are usually
concerned with arrivals over simple intervals, we will understand that
if \(N\) is indexed by a single point parameter, then it corresponds to
the following counting function: \(N(t):=N([0,t))-N((-\infty, t))\).

We represent a realisation of the time series by a (finite by
assumption) sequence of arrival times
\(\v{t}=\{\dots,t_1, t_2,\dots,t_N(T)\}.\) We will be concerned with
evaluating its likelihood at the arrivals in time \([0,T]\) We define
\(\v{t}':=\v{t}\cap [0,T]\) and assume \(\left|\v{t}'\right|<\infty.\)

@@

TODO: additional criteria are required for finite intensity process.

    We use only the left-continuous filtration
\(\mathcal{F}(\{N(s)\}_{s < t})\) generated by the paths of the process
to time \(t\), which we save ink by writing here as
\(\mathcal{F}(\{N(s)\}_{s < t}) \equiv \cc{H}_t.\)

We call it a regular point process if we may characterise the times of
the jumps of the process by an associated \emph{intensity} process
\(\lambda\), such that \(\lambda\) is predictable with respect to
\(\cc{H}_t\) i.e. \cite{daley_introduction_2003}

\[\begin{aligned}\lambda(t|\cc{H}_t)&:=\lim_{h\to 0} \frac{1}{h}\left(1-\bb P\left[N(t+h)=N(t)|\cc{H}_t\right] \right)\\
&=\lim_{h\to 0} \frac{1}{h}\left(\bb E \left[N([t,t+h))\right]\right).\end{aligned}\]

We will require that \(\bb E\lambda^*<\infty.\)

Such regular point processes generalise the homogeneous Poisson rate in
the sense that if we choose constant
\(\lambda(t|\cc{H}_t)\equiv \lambda_0,\) then it is the intensity of a
homogeneous Poisson rate with fixed intensity \(\lambda_0\), such that a
process \(N'\) with this intensity is Poisson,
\(N'(t)-N'(s)\sim\oo{Pois}(\lambda_0(t-s))\) for \(t\geq s \geq 0\).
More generally, we define \[
\Lambda(s,t) = \int_s^t\lambda(u,\cc{H}_u) \dd u.
\]

Then \[N(t)-N(s)\sim\oo{Pois}(\Lambda(s,t)),\] i.e. the pmf f
\(k=N(t)-N(s)\) is given

\[\bb P(N(t)-N(s)=k)=\frac {\Lambda^k(s,t)e^{-\Lambda(s,t) }}{k!},\] and
moreover, the increments of the process over disjoint intervals of the
index space are independent.

(TODO: not quite right - statement needs the ordering of the filtration
to be true)

    Considering each increment of the process , we have an expression for
the density of a given arrival given the process history
\[\begin{aligned}
f(t, N(t_i)) &= \frac{\dd}{\dd t}\bb P(t_{i+1}<t) \\
&= \frac{\dd}{\dd t} \bb P(N(t)=N(t_{i}))  \\
&= \frac{\dd}{\dd t} \frac {\Lambda^0(t_i,t)e^{-\Lambda(t_i,t) }}{0!}\\
&= \frac{\dd}{\dd t} e^{-\Lambda(t_i,t)} \\
&= \left(\frac{\dd}{\dd t} -\Lambda(t_i,t)\right) e^{-\Lambda(t_i,t)} \\
&= -\lambda(t, \cc H(t)) e^{-\Lambda(t_i,t)} 
\end{aligned}\]

    then we may write the density of \(f(t_{i+1}),\)

\[
f(t_{i+1}, N(t_i)) = -\lambda(t_{i+1}, \cc H(t_{i+1})) e^{-\Lambda(t_i,t_{i+1})} 
\] and the likelihood of the entire realisation as the product of all
conditional densities

\[
f(t_1, t_2, \dots t_{N(T)}) = \prod_{i=0}^{N(T)} \left(-\lambda(t_{i+1}, \cc H(t_{i+1})) e^{-\Lambda(t_i,t_{i+1})} \right)
\]

TODO term for last interval \((t_{N(T)},T],\) handle \(i=0.\)

    This is, as usual, more conveniently managed as a log likelihood

\[\begin{aligned}
\ell(\v t) &:= \log f(t_1, t_2, \dots t_{N(T)})\\
& = \sum_{i=0}^{N(T)}\left( -\log \lambda(t_{i+1}, \cc H(t_{i+1})) -\Lambda(t_i,t_{i+1})\right)\\
&=  \left(\sum_{i=1}^{N(T)} -\log \lambda(t_{i}, \cc H(t_{i}))\right) -\int_0^T \lambda(t, \cc H(t)
\end{aligned}\]

TODO missing minus sign for summation term.

@@

\[
\begin{aligned}
\lambda(t|\cc H_t;\mu,\phi) &= \mu(t) + \phi * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi(t-s)\dd N(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi(t-s)\dd N(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi(t-t_i)
\end{aligned}
\]

For brevity, we write
\(\lambda(\mathcal{F}(N(t));\mu,\phi)=\lambda^*(t)\).

    Using that \(N(t)=N(t_i)\ \forall t\lt t_{i+1}\) we can now write the
joint occurrence density as \cite{rubin_regular_1972}.
\[\begin{aligned}L(N;\mu,\phi) &=-\int_0^T\lambda(t|\cc H_t) \dd t + \int_0^T\log \lambda(t|\cc H_t) \dd N_t \\ 
    &=-\int_0^T\lambda(t|\cc H_t)\dd t + \int_0^T\log \lambda(t|\cc H_t) \dd N_t \\
    &=-\int_0^T\lambda(t|\cc H_t)\dd t + \sum_{t\in\v{t}'}\log \lambda(t|\cc H_t) 
\end{aligned}\]

    Substituting in the Hawkes intensity formula in particular gives us an
expression for the joint likelihood.

    Hereafter we define \(\lambda^*(t):=\lambda(t|\cc{H}_t)\) for brevity.

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_18_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_34_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_34_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_34_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

\texttt{\color{outcolor}Out[{\color{outcolor}7}]:}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_36_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

\texttt{\color{outcolor}Out[{\color{outcolor}8}]:}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_38_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

    This is suggestive of interpreting \(\lambda\) as a hazard rate, as
indeed it is. Considering the density function \(f\) and cumulative
density \(F\) of the time of the next event, TODO show this? start from
this?

\[\begin{aligned}
\lambda(t)&=\frac{f(t|\cc{H}_t)}{1-F(t|\cc{H}_t)}\\
&=\frac{f(t|\cc{H}_t)}{1-F(t|\cc{H}_t)}\\
\end{aligned}\]

    Rearranging the intensity, we note that \[\begin{aligned}
\frac{\dd}{\dd t} \bb P\left[N(t)=N(s)|\cc{H}_t\right] &= \lambda(t|\cc{H}_t) \\
&= 
\end{aligned}\] for \(t\geq s \geq 0\)

\[\begin{aligned}
\bb P\left[N(t)=N(s)|\cc{H}_s\right]\propto\exp\left[-\int_{s}^{t}\lambda^*(u) \dd u\right]
.\end{aligned}\]
